# IL Financial Contributions to 2016 Presidential Campaigns
========================================================

## Abstract

This is an exploration data analysis on IL 2016 US presidential campaign
contribution.

The dataset can be found [here](http://classic.fec.gov/disclosurep/PDownload.do)

Here is the list of variables contained in this file:

* cmte_id
* cand_id           
* cand_nm
* contbr_nm         
* contbr_city       
* contbr_st        
* contbr_zip        
* contbr_employer
* contbr_occupation
* contb_receipt_amt
* contb_receipt_dt
* receipt_desc"     
* memo_cd           
* memo_text         
* form_tp        
* file_num          
* tran_id
* election_tp

When I first get this dataset, and scan through it. Here are some questions I \
immediately have. I will try to answer these questions during the exploration. \
I will see if there are additional questions I can answer when I play around \
with the data.

* Which candidate receives the most money?
* Which candidate has the most supporters?
* Who donates the highest amount of money?
* What do the donors do?
* How do these donors donate?

## Setup

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(dplyr)
library(readr)
 
```

## Initial Data Exploration

```{r echo=FALSE, warning=FALSE, Load_the_Data}

il <- read_csv('contribution-data-IL.csv')

dim(il)
str(il)
names(il)

```

This is a big dataset, with 250k rows of data, and 18 variables. 

## Univariate Plots Section

### Receipt Amount Info

Let's take a quick look at the summary of this information.

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(il$contb_receipt_amt)

```

As you can seen in the summary. There are a lot of outliers in this dataset. \
Most donation is between 0 to 100. Unexpectedly, there are some negative \
donations as well. Details cannot be shown clearly in the orginal histogram. \
Therefore, I am going to add some transformation to this plot.

```{r echo=FALSE, warning=FALSE, Univariate_Plots1}

ggplot(data=il, aes(x=contb_receipt_amt)) +
    geom_histogram(bins=50) +
    scale_x_log10() +
    xlab('contribution amount') +
    ggtitle('Contribution Amount Distribution')

```

We can see the trends of this dataset more clearly right now. However, there \
is a potential issue. Since the dataset contains several negative \
contributions, which doesnt' work well with log10. In my analysis, I will omit \
these negative amount since they are refund when the contribution exceeds the [limits](https://www.fec.gov/help-candidates-and-committees/candidate-taking-receipts/contribution-limits-candidates/). They are not very helpful for my analysis. \
Based on the warning messages, around 3k rows are removed, which is around 1% \
of the total dataset. 

### Candidate Info

```{r echo=FALSE, warning=FALSE, Univariate_Plots2}

table(il$cand_nm)

ggplot(aes(x = cand_nm), data = il) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('candidate') +
    ylab('# of Donations') +
    ggtitle('# of Donation per Candidate')

```

We can clearly see that Hillary Clinton has the highest amount of supporters. \
Sanders is on the second spot, roughly 1/2 of Clinton. Donald Trump is the third
one, rougly 1/5 of Clinton. 

### Contributor Occupation

The contributor's occupation information is not clean. Based on a quick summary
of this part of data, the name of occupation is not consistent. There are a lot
of occupation with only 1 contributor. Therefore, to better understand it, I
will only focus on the top 10 occupations.

```{r echo=FALSE, warning=FALSE, Univariate_Plots3}
occupation_group <- group_by(il, contbr_occupation)

il_by_occupation <- summarize(occupation_group,
                              sum_occupation = sum(contb_receipt_amt),
                              mean_occupation = mean(contb_receipt_amt),
                              n = n())

il_by_occupation <- subset(il_by_occupation,
                           contbr_occupation != "INFORMATION REQUESTED")

il_by_occupation <- head(arrange(il_by_occupation, desc(n)), n = 10)

il_by_occupation$contbr_occupation

il_by_occupation$contbr_occupation <-
    ordered(il_by_occupation$contbr_occupation,
            levels = c('RETIRED', "NOT EMPLOYED", "ATTORNEY", "TEACHER",
                       "PROFESSOR", "PHYSICIAN", "HOMEMAKER", "CONSULTANT",
                       "SALES", "LAWYER"))

ggplot(data=il_by_occupation, aes(x=contbr_occupation, y=n)) +
    geom_bar(stat = 'identity') +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    xlab('Occupation') +
    ylab('Count') +
    ggtitle('# of Contribution per Occupation')

```

Based on a quick summary for the contribution information, we can see that
retired people are more involved in the presidential campaign, following by
non-employed people. 

### Party

The dataset doesn't have the party information of all the candidates. I will
create the party variable based on my understanding of each candidate. 

```{r echo=FALSE, warning=FALSE, Univariate_Plots4}

unique(il$cand_nm)

democrat <- c("Clinton, Hillary Rodham", "Sanders, Bernard", "O'Malley,
              Martin Joseph", "Lessig, Lawrence", "Webb, James Henry Jr.")
others <- c("Johnson, Gary", "McMullin, Evan", "Stein, Jill")
il$party <- ifelse(il$cand_nm %in% democrat, "democrat",
                   ifelse(il$cand_nm %in% others, "others", "republican"))

party_group <- group_by(il, party)
il_contr_by_party <- summarize(party_group,
                               sum_party = sum(contb_receipt_amt),
                               number_of_candidate = length(unique(cand_nm)), 
                               mean_party = sum_party/number_of_candidate, 
                               n = n()) 
                              
il_contr_by_party
il_contr_by_party$party <- ordered(il_contr_by_party$party, 
                                   levels = c('democrat', 'republican', 'others'))

ggplot(data = il_contr_by_party, aes(x = party, y = n, fill = party)) +
  geom_bar(stat = 'identity') +
  geom_text(stat = 'identity', aes(label = n),
            data = il_contr_by_party, vjust = -0.4) +
  xlab('Party') +
  ylab('# of Contributions') +
  ggtitle('# of Contributions by Party') +
  scale_fill_manual(values = c('dark blue', 'red', 'green'))

```


## Univariate Analysis

### What is the structure of your dataset?

This dataset has 250,411 observations, and 18 variables (features). The detail
list of the all the variables can be found in the Abstract section.
The features I am most interested in this datasets are the candidates and the
amount they received.


### What is/are the main feature(s) of interest in your dataset?

I am mainly interested in the contribution amount, who is contributing and how
much each candidate received. I will try to answer the questions in the abstract
section.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The occupation is the feature that can help support my investigation. As the
plots showing above, the average of contribution amount varies a lot among
different occupations.

### Did you create any new variables from existing variables in the dataset?

There is only candidate name and id information instead of party information,
which is an important factor when analyzing presidential campaign. Therefore,
I created a "party" variable using the candidate name.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

As mentioned above, I omitted the negative contribution, which is the refund
amount that exceeds the $2,700 limitation.

## Bivariate Plots Section

### Occupation & Contribution Amount

```{r echo=FALSE, warning=FALSE, Bivariate_Plots1}
il_by_occupation[with(il_by_occupation, order(mean_occupation)), ]

il_by_occupation$contbr_occupation <-
    ordered(il_by_occupation$contbr_occupation,
            levels = 
                c('ATTORNEY', 'LAWYER', 'HOMEMAKER', 'PHYSICIAN', 'CONSULTANT',
                'RETIRED', "PROFESSOR", 'SALES', "TEACHER", "NOT EMPLOYED"))

ggplot(data=il_by_occupation, aes(x=contbr_occupation, y=mean_occupation)) +
    geom_bar(stat = 'identity') +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    xlab('Occupation') +
    ylab('Average Contribution Amount') +
    ggtitle('Average Contribution per Occupation')
```

As discussed above, retired and non-employed people are more involved in the presidential campaign. However, from an average contribution perspective,
attorney has the highest average contribution, following by lawyer and
homemaker, whose average contribution is almost the same. Non-employed people
have the lowest average contribution. 

```{r echo=FALSE, warning=FALSE, Bivariate_Plots2}

il_by_occupation <- head(arrange(il_by_occupation, desc(sum_occupation)),
                         n = 10)
il_by_occupation

il_by_occupation$contbr_occupation <-
    ordered(il_by_occupation$contbr_occupation,
            levels = 
                c('RETIRED', 'ATTORNEY', 'HOMEMAKER', 'NOT EMPLOYED',
                  'PHYSICIAN', 'LAWYER', 'CONSULTANT', "PROFESSOR", "TEACHER",
                  'SALES'))

ggplot(data=il_by_occupation, aes(x=contbr_occupation, y=sum_occupation)) +
    geom_bar(stat = 'identity') +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    xlab('Occupation') +
    ylab('Average Contribution Amount') +
    ggtitle('Average Contribution per Occupation')

```

### Party & Contribution Amount

We have seen the # of contribution per party in the previous exploration. Let's
take a look at the contribution amount per party right now.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots3}

ggplot(data = il_contr_by_party, 
       aes(x = party, y = sum_party/1000, fill = party)) +
    geom_bar(stat = 'identity') +
    geom_text(data = il_contr_by_party, stat = 'identity',
              aes(label = round(sum_party/1000)),
              vjust = -0.5) +
    xlab('Party') +
    ylab('Contribution Amount (k)') +
    ggtitle('Total Contribution Amount per Party') +
    scale_fill_manual(values = c('dark blue', 'red', 'green'))

subset(il_contr_by_party, party == 'democrat')$sum_party /
    sum(il_contr_by_party$sum_party)

```

There is no surprise here. Democracy still receives twice as much as republican
received. Other parties receive very small amounts. Domocracy recevied almost \
70% of the total contribution.

```{r echo=FALSE, warning=FALSE, Bivariate_Plots4}

ggplot(data = il_contr_by_party, aes(x = party, y =  mean_party/1000, 
                                     fill = party)) +
    geom_bar(stat = 'identity') +
    xlab('Party') +
    ylab('Contribution Amount (k)') +
    ggtitle('Average Contribution per Party') +
    scale_fill_manual(values = c('dark blue', 'red', 'green'))

subset(il_contr_by_party, party == 'democrat')$mean_party /
    subset(il_contr_by_party, party == 'republican')$mean_party

```

There are even more differences here. Democracy received more than 9 times of
the contribution as much as republican received. 

### Candidate & Contribution Amount

```{r echo=FALSE, warning=FALSE, Bivariate_Plots5}
sort(with(il, by(contb_receipt_amt, cand_nm, sum)))

ggplot(data = il, aes(x = cand_nm, y = contb_receipt_amt/1000)) +
    geom_bar(stat = 'identity') + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Candidate') +
    ylab('Total Contribution Amount (k)') +
    ggtitle('Total Contribution per Candidate')

sum(subset(il, cand_nm == 'Clinton, Hillary Rodham')$contb_receipt_amt) /
    sum(il$contb_receipt_amt)

```

Unsurprisingly, Hillary Clinton received the highest contribution amount, \
60% of the total contribution.

## Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

* Although in terms of total amount of people, retired people and unemployed \
people are more involved in the presidential campaign, in terms of average \
contribution, attorney, laywer and homemakers are the lead. 
* In Illinois, Democracy received 68% of the total contribution, with only 5 \
Democracy candidates and 17 Republican candidates.
* The average amount received by Democracy is more than 9 times of what \
Republican received.
* Hillary Clinton alone received 57% of the total contribution.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Homemakers have the average contribution almost the same with lawyers. They \
have the 3rd highest average contribution in Illinois.

I did really know there is a big difference between attorney and lawyer. \
However, just from the contribution perspective, attorneys are more involved, \
and contribute more than lawyers.

### What was the strongest relationship you found?

In Illinois, people support Democracy more, especially Hillary Cliton. 

## Multivariate Plots Section

```{r echo=FALSE, warning=FALSE, Multivariate_Plots}

top_candidate <- c('Clinton, Hillary Rodham', 'Sanders, Bernard',
                   'Trump, Donald J.')

il_top_candidate <- il %>%
  filter(cand_nm %in% top_candidate) %>%
  group_by(cand_nm, contb_receipt_dt) %>%
  summarize(n = n(), total = sum(contb_receipt_amt))

il_top_candidate$contb_receipt_dt <-
    as.Date(il_top_candidate$contb_receipt_dt, "%d-%b-%y")

il_top_candidate[order(as.Date(
    il_top_candidate$contb_receipt_dt, format="%m/%y")),]


ggplot(data = il_top_candidate,
    aes(x = contb_receipt_dt, y = total, color = cand_nm)) +
        geom_jitter() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Date') +
    ylab('Total Contribution') +
    ggtitle('Total Contribution per Candidate per Month')

```

We can see, from this plot, that Hiliary Clinton leads the total contribution \
amount. Bernard Sanders has more contribution from 2015 to 2016. Donald Trump \
has few contributions at the beginning, but starts to get more since 2016.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

* Hiliary Clinton leads the total contribution amount all the time. She \
received more contributions towards the end of the campaign.
* Bernard Sanders has more contribution from 2015 to 2016, but almost none at \
the end of the campaign.
* Donald Trump has few contributions at the beginning, but starts to get more \
since 2016.

### Were there any interesting or surprising interactions between features?

For a very short period of time, Bernard Sanders received more contributions \
than Hillary Clinton.

------

## Final Plots and Summary

### Plot One: Most people contribution $100 or less
```{r echo=FALSE, warning=FALSE, Plot_One}

ggplot(data=il, aes(x=contb_receipt_amt)) +
    geom_histogram(bins=50) +
    scale_x_log10() +
    xlab('contribution amount') +
    ggtitle('Contribution Amount Distribution')
```

### Description One

I am not very familiar with the presidential campaign. When I first saw the \
dataset, I thought I could find people who was very crazy and would donate \
thousands of dollars. To my surprise, most people donate $100 or less. Later, \
I noticed that there are acutally $2,700 contribution limitation.

### Plot Two: Attorney has the highest average contribution
```{r echo=FALSE, warning=FALSE, Plot_Two}
il_by_occupation[with(il_by_occupation, order(mean_occupation)), ]

il_by_occupation$contbr_occupation <-
    ordered(il_by_occupation$contbr_occupation,
            levels = 
                c('ATTORNEY', 'LAWYER', 'HOMEMAKER', 'PHYSICIAN', 'CONSULTANT',
                'RETIRED', "PROFESSOR", 'SALES', "TEACHER", "NOT EMPLOYED"))

ggplot(data=il_by_occupation, aes(x=contbr_occupation, y=mean_occupation)) +
    geom_bar(stat = 'identity') +
    theme(axis.text.x = element_text(angle=90, hjust=1)) +
    xlab('Occupation') +
    ylab('Average Contribution Amount') +
    ggtitle('Average Contribution per Occupation')
```

### Description Two

As discussed above, retired and non-employed people are more involved in the presidential campaign. However, from an average contribution perspective,
attorney has the highest average contribution, following by lawyer and
homemaker, whose average contribution is almost the same. Non-employed people
have the lowest average contribution. 

### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Three}

top_candidate <- c('Clinton, Hillary Rodham', 'Sanders, Bernard',
                   'Trump, Donald J.')

il_top_candidate <- il %>%
  filter(cand_nm %in% top_candidate) %>%
  group_by(cand_nm, contb_receipt_dt) %>%
  summarize(n = n(), total = sum(contb_receipt_amt))

il_top_candidate$contb_receipt_dt <-
    as.Date(il_top_candidate$contb_receipt_dt, "%d-%b-%y")

il_top_candidate[order(as.Date(
    il_top_candidate$contb_receipt_dt, format="%m/%y")),]


ggplot(data = il_top_candidate,
    aes(x = contb_receipt_dt, y = total, color = cand_nm)) +
        geom_jitter() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Date') +
    ylab('Total Contribution') +
    ggtitle('Total Contribution per Candidate per Month')
```

### Description Three

We can see, from this plot, that Hiliary Clinton leads the total contribution \
amount. Bernard Sanders has more contribution from 2015 to 2016. Donald Trump \
has few contributions at the beginning, but starts to get more since 2016.

------

# Reflection

## Challenages & Struggles

* I actually had some issues reading the dataset at the beginning. I used \
`read.csv()` at first, but it keeps generating errors. I used \
`row.names = NULL`, but it generated a row named `NULL` and messed up with \
the rest of the dataframe. After some research, I used `read_csv()`, which \
doesn't have any issues. I will use `read_csv()` going forward.
* I have some struggles with time seires variables in R. In the original \
dataset, the date information was treaed as characters instead of Date type. \
It took me a while to clean it up to draw the time seris plot.

## Future Work

This dataset only contains information from Illinois, which is known as the \
state supports Democracy. I am interesting to explore and analyze the campaign \
information in other state as well, which could be a total different story.