# IL Financial Contributions to 2016 Presidential Campaigns
========================================================

## Abstract

This is an exploration data analysis on IL 2016 US presidential campaign
contribution.

The dataset can be found [here](http://classic.fec.gov/disclosurep/PDownload.do)

Here is the list of variables contained in this file:

* cmte_id
* cand_id           
* cand_nm
* contbr_nm         
* contbr_city       
* contbr_st        
* contbr_zip        
* contbr_employer
* contbr_occupation
* contb_receipt_amt
* contb_receipt_dt
* receipt_desc"     
* memo_cd           
* memo_text         
* form_tp        
* file_num          
* tran_id
* election_tp

When I first get this dataset, and scan through it. Here are some questions I immediately have. I will try to answer these questions during the exploration.
I will see if there are additional questions I can answer when I play around 
with the data.

* Which candidate receives the most money?
* Which candidate has the most supporters?
* Who donates the highest amount of money?
* What do the donors do?
* How do these donors donate?

## Setup

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(dplyr)
library(readr)
 
```

## Initial Data Exploration

```{r echo=FALSE, Load_the_Data}

il <- read_csv('contribution-data-IL.csv')

dim(il)
str(il)
names(il)

```

This is a big dataset, with 250k rows of data, and 18 variables. 

## Univariate Plots Section

### Receipt Amount Info

Let's take a quick look at the summary of this information.

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(il$contb_receipt_amt)

```

As you can seen in the summary. There are a lot of outliers in this dataset.
Most donation is between 0 to 100. Unexpectedly, there are some negative
donations as well. Details cannot be shown clearly in the orginal histogram.
Therefore, I am going to add some transformation to this plot.

```{r echo=FALSE, Univariate_Plots}

ggplot(data=il, aes(x=contb_receipt_amt)) +
    geom_histogram(bins=50) +
    scale_x_log10()

```

We can see the trends of this dataset more clearly right now. However, there is
a potential issue. Since the dataset contains several negative contributions, 
which doesnt' work well with log10. In my analysis, I will omit these negative
amount since they are refund when the contribution exceeds the [limits](https://www.fec.gov/help-candidates-and-committees/candidate-taking-receipts/contribution-limits-candidates/). They are not very helpful for my analysis.
Based on the warning messages, around 3k rows are removed, which is around 1%
of the total dataset. 

### Candidate Info

```{r echo=FALSE, Univariate_Plots}

table(il$cand_nm)

ggplot(aes(x = cand_nm), data = il) +
    geom_bar() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('candidate') +
    ylab('# of Donations')

```

We can clearly see that Hillary Clinton has the highest amount of supporters.
Sanders is on the second spot, roughly 1/2 of Clinton. Donald Trump is the third
one, rougly 1/5 of Clinton. 

### Contributor Occupation

The contributor's occupation information is not clean. Based on a quick summary
of this part of data, the name of occupation is not consistent. There are a lot
of occupation with only 1 contributor. Therefore, to better understand it, I
will only focus on the top 10 occupations.

```{r echo=FALSE, Univariate_Plots}
occupation_group <- group_by(il, contbr_occupation)

il_by_occupation <- summarize(occupation_group,
                              sum_occupation = sum(contb_receipt_amt),
                              mean_occupation = mean(contb_receipt_amt),
                              n = n())

il_by_occupation <- subset(il_by_occupation,
                           contbr_occupation != "INFORMATION REQUESTED")

il_by_occupation <- head(arrange(il_by_occupation, desc(n)), n = 10)

il_by_occupation$contbr_occupation

il_by_occupation$contbr_occupation <-
    ordered(il_by_occupation$contbr_occupation,
            levels = c('RETIRED', "NOT EMPLOYED", "ATTORNEY", "TEACHER",
                       "PROFESSOR", "PHYSICIAN", "HOMEMAKER", "CONSULTANT",
                       "SALES", "LAWYER"))

ggplot(data=il_by_occupation, aes(x=contbr_occupation, y=n)) +
    geom_bar(stat = 'identity') +
    theme(axis.text.x = element_text(angle=90, hjust=1))

il_by_occupation[with(il_by_occupation, order(mean_occupation)), ]

il_by_occupation$contbr_occupation <-
    ordered(il_by_occupation$contbr_occupation,
            levels = c('ATTORNEY', 'LAWYER', 'HOMEMAKER', 'PHYSICIAN', 'CONSULTANT',
                       'RETIRED', "PROFESSOR", 'SALES', "TEACHER", "NOT EMPLOYED"))

ggplot(data=il_by_occupation, aes(x=contbr_occupation, y=mean_occupation)) +
    geom_bar(stat = 'identity') +
    theme(axis.text.x = element_text(angle=90, hjust=1))
```

Based on a quick summary for the contribution information, we can see that retired people are more involved in the presidential campaign, following by non-employed people. However, from an average contribution perspective, attorney has the highest average contribution, following by lawyer and homemaker, whose average contribution is almost the same. Non-employed people have the lowest average contribution. 

## Univariate Analysis

### What is the structure of your dataset?

This dataset has 250,411 observations, and 18 variables (features). 

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?



### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


## Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

## Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


## Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

## Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!